---
title: "Hw3-#1"
author: "Jiyoon Clover Jeong"
date: "10/13/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\paragraph{\#3 -- Ants}
\Solution
First, it should be noted that the data contained a row with an \texttt{NA}.
Consequently, that row was excluded for the entirety of the analysis.
Furthermore, that 'Distance' only took on 5 unique values suggested that we 
consider the variable as categorical, rather than continuous. Hence, the 
proceeding analysis was begun with 3 categorical predictors ('Colony',
'Distance', and 'Size.class') and 1 continuous predictor ('Headwidth').

The simple EDA performed in figures 9 and 10 indicate that worker $Headwidth$
and $Size.class$ distribution do vary somewhat between colonies. In particular,
Colony 4 and Colony 1 have a disproportionate amount of workers of with
$Size.class = 40-43$. Colony 6 seems to have most of its workers in $Size.class$
greater than 35. Worker $Mass$ is the least variable (across colonies). 

In fitting the model $Mass \sim Colony + Distance + Size.class + Headwidth$,
a few unexpected results occured. Namely, the the coefficients on the dummy
variables for the 'Class.size' were negative (and very insignificant). Given
that most biological models would imply a direct relationship between an 
animal's size and its mass, these coefficients put the validity of this model
in to question. Since it doesn't make sense to transform categorical indicators
there were only two ways to fix this "problem":
\begin{enumerate}
  \item Drop 'Class.size' (since its coefficients had incredibly high 
  $p$-values), or
  \item Consider respecifying 'Headwidth' (the only numeric variable in the
  model).
\end{enumerate}
Given there's a very good reason to keep 'Class.size' in the model, option \#1
is not reasonable. For option \#2, residual, AV, and CPR plots were used to
investigate if any unexplained curvature is present (see figure
\ref{fig:q3RegAnalysisPlots}) from 'Headwidth'. Surprisingly, adding a 
quadratic transformation of 'Headwidth' removed any structure between the
regression model's residuals and 'Headwidth'. Furthermore, the \textit{adjusted}
$R^2$ value for the model that adds $Headwidth^2$ to the previous "full model"
takes on a 0.03\% increase.


```{r}

<<loadData>>=
setwd(dir="~/Dropbox/Berkeley/STAT151A/")
data <- read.table(file="HW_solutions/HW3/hw3_ants.txt", header=TRUE,
                      na.strings=c("NA", "\x80"),
                      colClasses=c(rep.int("factor",2),
                                   rep.int("numeric",3),
                                   "factor"))
@

<<q3, dependson=c("loadData")>>=
ants.df <- data[apply(data, 1, function(row) all(!is.na(row))), ]
ants.df <- ants.df[, -5]
@

```


```{r}

<<q3ColonyPlots, dependson=c("q3"), fig.cap="Basic, colony-level, EDA of variables in \\texttt{ants} dataset.">>=
basePlot <- ggplot(data=ants.df)

basePlot +
  geom_histogram(aes(x=Distance, y=..count.., fill=Colony),
                 position="dodge") +
  labs(y=NULL, title="Histogram of Distance for the 6 Colonies") +
  theme_bw()

basePlot +
  geom_boxplot(aes(y=Mass, x=Colony)) +
  geom_point(aes(y=Mass, x=Colony, color=Colony), shape=1) +
  theme_bw() +
  labs(x=NULL, title="Distribution of worker mass for the 6 Colonies")

basePlot +
  geom_boxplot(aes(y=Headwidth, x=Colony)) +
  geom_point(aes(y=Headwidth, x=Colony, color=Colony), shape=1,
             position=position_jitter(width=0.05)) +
  theme_bw() +
  labs(x=NULL, title="Distribution of worker headwidth for the 6 Colonies")

basePlot +
  geom_histogram(aes(x=Size.class, y=..count.., fill=Colony),
                 position="dodge") +
  labs(y=NULL,
       title="Histograms of size classes of the workers in the 6 Colonies") +
  theme_bw()
@


```



```{r}
<<q3ColonyPairsPlots, dependson=c("q3"), fig.cap="Colony-level pairs plots.">>=
tmp <- sapply(levels(ants.df$Colony), function(lvl) {
  title <- paste0("Pairs plot for Colony ", lvl)
  pairs(subset(ants.df, Colony == lvl, select=-Colony),
        panel=panel.smooth, main=title)
})
@


```




```{r}

<<q3RegAnalysis, dependson=c("q3"), cache=FALSE>>=
fullModel <- lm(Mass ~ 0 + ., data=ants.df)

partialModel <- lm(Mass ~ 0 + Colony + Size.class + Distance, data=ants.df)

# check that adding headwidth is smart
partialModel2 <- lm(Headwidth ~ 0 + Colony + Size.class + Distance,
                    data=ants.df)

partialModel3 <- lm(Headwidth + I(Headwidth^2) ~ 0 + .,
                    data=ants.df[, -3])

fullModel2 <- lm(Mass ~ 0 + . + I(Headwidth^2), data=ants.df)
@
<<q3RegAnalysisTab1, dependson=c("q3RegAnalysis"), results='asis'>>=
fullModelAdjR2 <- round(x=summary(fullModel)$adj.r.squared,
                        digits=4)
cap <- paste0("\\texttt{summary} for model including all variables in
             ants data. Adjusted $R^2$ for this model is ", fullModelAdjR2)
print(xtable(fullModel, caption=cap))

parModelAdjR2 <- round(summary(partialModel)$adj.r.squared, 4)
cap <- paste0("\\texttt{summary} for model including  only including
             'Colony', 'Distance', and 'Size.class'. Adjusted $R^2$ for this
             model is ", parModelAdjR2, ".Note that all variables are
             signficant, here.")
print(xtable(partialModel, caption=cap)) # all variables significant with

fullModel2AdjR2 <- round(summary(fullModel2)$adj.r.squared, 4)
cap <- paste0("\\texttt{summary} for model including all variables in ",
             "ants data as well as 'Headwidth$^2$'. Adjusted $R^2$ for this ",
             "model is ", fullModel2AdjR2, ", which is ",
             round(100*(fullModel2AdjR2 - fullModelAdjR2)/fullModelAdjR2,3),
             "\\% more than the previous full model. Note that, we're in a ",
             "position where most of our variables are significant and the ",
             "relationships between 'Mass' and 'Size.class' make sense. (I.e.,",
             " there's a positive, trend between the latter and the former.)")

@

```


The summary of the \texttt{lm} call is below. See tables 1 and 2 for the 
print-outs of $Mass ~ Size.class + Distance + Headwidth + Colony$ and 
$Mass ~ Size.class + Distance + Colony$, respectively. 





```{r}

<<q3RegAnalysisTab2, dependson=c("q3RegAnalysis", "q3RegAnalysisTab1")>>=
summary(fullModel2, caption=cap)
@

```

Towards the conclusions we might draw from this model:
\begin{enumerate}
  \item Inside the general distribution of ant mass, the colony-level
  contribution does not vary greatly; the mean and standard deviation of the 
  coefficients are 
  \Sexpr{coeffs <- c(162.4777,161.4726,166.4157 ,161.6677,165.7055,161.7623);
  mean(coeffs)} and \Sexpr{sd(coeffs)}, respectively.
  \item There's a statistically significant \textit{negative} relationship
  between $Mass$ and $Distance$. Given that $Mass$ also includes any food
  carried by the specimen, this may lead one to conclude that ants,
  in general, employ the "energy conservative" strategy. That is, our full
  model suggests that, all other factors held constant, moving from one
  distance level to a higher level will yield a decrease in mass. However,
  colony-level analysis shows that this isn't the full picture (see figure
  \ref{fig:q3ColonyRegression}). In particular, only Colony 1 had a model whose
  'Distance' coefficients were negative (see table 3). All other Colonies had 
  models with positive 'Distance' coefficients. That being said, all Colony
  models had \textit{adjusted} $R^2$ values greater than than $0.98$, giving us
  some assurance that these models are capturing a lot of the underlying
  variability in $Mass$. This discrepancy amongst the colonies should lead us
  to conclude that each colony employs its own strategy. Namely, Colony 1
  is ``energy conservative'' while the other colonies are 
  ``worker conservative''. 
\end{enumerate}



```{r}

<<q3RegAnalysisPlots, fig.cap="The residual plot in the top left indicates some unexplained curvature that the AV and CPR plots corroborate. After modeling 'Mass' $\\sim$ 'Colony' + 'Size.class' + 'Distance' + 'Headwidth' + 'Headwidth$^{2}$', the residuals in the bottom right plot don't seem to indicate any extra curvature.", dependson=c("q3RegAnalysis")>>=

ggplot(data=data.frame(Y=fullModel$residuals,
                       X=ants.df$Headwidth),
       aes(x=X,y=Y)) +
  geom_point(shape=1) +
  stat_smooth(method="loess", se=FALSE, color='red', lty=2) +
  stat_smooth(method="lm", se=FALSE, color='blue', alpha=0.65) +
  theme_bw() +
  labs(x="Headwidth",
       y="Full Model Residual's",
       title="Residuals versus Headwidth")

ggplot(data=data.frame(Y=partialModel$residuals, X=partialModel2$residuals),
       aes(x=X,y=Y)) +
  geom_point(shape=1) +
  stat_smooth(method="loess", se=FALSE, color='red', lty=2) +
  stat_smooth(method="lm", se=FALSE, color='blue') +
  theme_bw() +
  labs(x="Headwidth ~ Partial Model",
       y="Partial Model's residuals",
       title="AV-plot for Headwidth")

ggplot(data=data.frame(Y=fullModel$residuals +
                         fullModel$coefficients["Headwidth"] *
                         ants.df$Headwidth,
                       X=ants.df$Headwidth),
       aes(x=X,y=Y)) +
  geom_point(shape=1) +
  stat_smooth(method="loess", se=FALSE, color='red', lty=2) +
  stat_smooth(method="lm", se=FALSE, color='blue', alpha=0.65) +
  theme_bw() +
  labs(x="Headwidth",
       y="Headwidth's Component + Residual",
       title="CPR-plot for Headwidth")

fun <- function(x) {
  sum(fullModel2$coefficients[c("Headwidth", "I(Headwidth^2)")] * c(x,x^2))
}

ggplot(data=data.frame(Y=fullModel2$residuals,
                       X=ants.df$Headwidth),
       aes(x=X,y=Y)) +
  geom_point(shape=1) +
  stat_smooth(method="loess", se=FALSE, color='red', lty=2) +
  stat_smooth(method="lm", se=FALSE, color='blue') +
  theme_bw() +
  labs(x="Headwidth", y="Residuals",
       title="Residuals from model with Headwidth + Headwidth^2 vs. Headwidth")
@
<<q3ColonyRegression, dependson=c("q3"), results='asis', out.width="0.85\\textwidth", fig.cap="Least-squares coefficients for full models fit on the individual colonies.">>=
ants.df <- within(ants.df, Headwidth2 <- Headwidth^2)
tmp <- lapply(levels(ants.df$Colony), function(lvl) {
  model <- lm(Mass ~ 0 + ., data=subset(ants.df, Colony==lvl, select=-Colony))
  coefs <- coefficients(summary(model))[,c(1,4)]
  adjustedR2 <- round(summary(model)$adj.r.squared, 4)
  
  cap <- paste0("\\texttt{summary} for full model regression on Colony ",
                lvl, " with adjusted $R^2 = $", adjustedR2, ".")
  print(xtable(model, caption=cap))
  out <- matrix(coefs[,1], nrow=1)
  colnames(out) <- names(coefficients(model))
  out <- data.frame(Colony=lvl, out)
  cbind(melt(out, id.vars=c("Colony")), p=coefs[,2])
})

tmp <- do.call(rbind, tmp)
ggplot(data=tmp,
       aes(y=variable, x=value, color=Colony,
           shape=cut(p, breaks=c(0, 0.01, 0.05, 0.1, 1)))
       ) +
  geom_point(size=3) +
  scale_shape_discrete("p-value", solid=FALSE)+
  labs(y=NULL, x="Least-squares coefficient") +
  theme_bw()

```

% Notes: 
% 1. converted "Distance" to a categorical variable, since it had 5 unique
% levels.
% 2. removed rows with NA's
% 3. removed "headwidth..mm." column since it's a linear transformation of Headwith
% EDA comments:
% Class.Size = (XS, S, M, L, XL)
% Colony 6 is predominantly M, L, and XL ants
% Colonies 1 and 4 have an abnormally high number of L ants
% Full model (without interactions):
% There's a direct relationship between size and mass: meaning, as we step up
% from one class to the next class 


```{r}



```




```{r}



```





