---
title: "151AHW4"
author: "Jiyoon Clover Jeong"
date: "10/26/2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(SignifReg) #stepwise 
library(caret) #fold
library(leaps, warn.conflicts = T) #Best subset
library(stats) #extract   AIC or AIC
```



# Problem 2 - (a)

```{r}
bodyfat <- read.csv("/Users/cloverjiyoon/2017Fall/Stat 151A/Lab/Lab3/bodyfat.csv")

bodyfat$Density <- NULL

names(bodyfat)

```


## Backward elimination using the individual p-values.

```{r}
dat1 <- bodyfat
n <- ncol(dat1)-1


for(i in 1:n){
  fit <- summary(lm(bodyfat ~., data = dat1))
  maxval <- max(fit$coefficients[-1,4])
  maxindex <- which.max(fit$coefficients[-1,4])
  # deleting intercept, getting pval column from summary
  if(maxval > 0.05){
      dat1 <- dat1[,-(maxindex + 1)]
  }
  else
    break
}

model1 <- dat1

summary(lm(bodyfat ~., data = model1))


# Check with function in R
SignifReg(bodyfat ~., data = bodyfat, alpha = 0.05, direction = "backward", criterion = "p-value")
```


## Forward Selection using p-values.

```{r}

dat1 <- bodyfat
formula <- c("bodyfat ~ 1")
predictors <- c()
plist <- as.vector(names(dat1[,-1]))
plist



for(i in 1:n){
  
  minindex <- 0
  minval <- 10^3
  
  for(j in 1:length(plist)){
    
    fit <- summary(lm( paste(formula, plist[j], sep = "+"),
                       data = dat1))
    if(fit$coefficients[plist[j], 4] < minval){
      minval <- fit$coefficients[plist[j], 4]
      minindex <- j
    }
  }
    
    # deleting intercept, getting pval column from summary
    if(minval > 0.05){
        break
    }
  
    else{
      formula <- paste(formula, plist[minindex], sep ="+")
      
      recovered_minindex <- match(plist[minindex], names(bodyfat))
      plist <- plist[-minindex]
      # recover to original index from plist
      predictors <- append(predictors, recovered_minindex)
      
    }
}


plist

formula

predictors

model2 <- bodyfat[,c(1,predictors)]
  
summary(lm(bodyfat ~., data = model2))


SignifReg(bodyfat ~., data = bodyfat, alpha = 0.05, direction = "forward", criterion = "p-value")


```


## Adjusted R2

```{r}

formula <- c("bodyfat ~ 1")
predictors <- c()
plist <- as.vector(names(dat1[,-1]))
plist



for(i in 1:n){
 
  prevR2 <-  summary(lm( formula, data = dat1))$adj.r.squared
  maxindex <- 0
  maxval <- 0
  
  for(j in 1:length(plist)){
    
    fit <- summary(lm( paste(formula, plist[j], sep = "+"),
                       data = dat1))
    fit
    if(fit$adj.r.squared > maxval){
      maxval <- fit$adj.r.squared
      maxindex <- j
    }
  }
    
    # deleting intercept, getting pval column from summary
    if(maxval < prevR2){
        break
    }
  
    else{
      formula <- paste(formula, plist[maxindex], sep ="+")
      
      recovered_maxindex <- match(plist[maxindex], names(bodyfat))
      plist <- plist[-maxindex]
      # recover to original index from plist
      predictors <- append(predictors, recovered_maxindex)
      
    }
}


plist

formula

predictors

model3 <- bodyfat[,c(1,predictors)]
  
summary(lm(bodyfat ~., data = model3))


```


## AIC  (pick smallest)

```{r}


AIC <- c()


subsets <- summary(regsubsets(bodyfat~., bodyfat, nvmax = 13,  method ="forward"))

m <- nrow(bodyfat)

for(i in 1:n){
  
  index <- as.numeric(subsets$which[i,])
  index <- which(index %in% c(1))
  index
  dat1 <- bodyfat[, index]
  head(dat1)
  residual <- residuals(summary(lm(bodyfat ~., data = dat1)))
  RSS_m <- sum((residual)^2)
  RSS_m
  AIC[i] <- m * log(RSS_m / m) + 2 * (1+ (length(index) -1))
  AIC[i]
}

which.min(AIC)
AIC

index <- as.numeric(subsets$which[which.min(AIC),])
index <- which(index %in% c(1))
index

model4 <- bodyfat[, index]

summary(lm(bodyfat~., data = model4))


```


## BIC
```{r}



BIC <- c()


subsets <- summary(regsubsets(bodyfat~., bodyfat, nvmax = 13,  method ="forward"))

m <- nrow(bodyfat)

for(i in 1:n){
  
  index <- as.numeric(subsets$which[i,])
  index <- which(index %in% c(1))
  index
  dat1 <- bodyfat[, index]
  head(dat1)
  residual <- residuals(summary(lm(bodyfat ~., data = dat1)))
  RSS_m <- sum((residual)^2)
  RSS_m
  BIC[i] <- m * log(RSS_m / m) + log(m) * (1+ (length(index) -1))
  BIC[i]
}

which.min(BIC)
BIC

index <- as.numeric(subsets$which[which.min(BIC),])
index <- which(index %in% c(1))
index

model5 <- bodyfat[, index]

summary(lm(bodyfat~., data = model5))

```


## Mallow's $C_p$

```{r}


C_p <- c()

# 
# subsets <- summary(regsubsets(bodyfat~., bodyfat, nvmax = 13,  method ="forward"))
# 
# m <- nrow(bodyfat)

for(i in 1:n){
  
  index <- as.numeric(subsets$which[i,])
  index <- which(index %in% c(1))
  index
  dat1 <- bodyfat[, index]
  head(dat1)
  residual <- residuals(summary(lm(bodyfat ~., data = dat1)))
  RSS_m <- sum((residual)^2)
  RSS_m
  
  residual <- residuals(summary(lm(bodyfat ~., data = bodyfat)))
  RSS <- sum(residual^2)
  
  sigmasq <- RSS / (m - (ncol(bodyfat) - 1) - 1)
  
  C_p[i] <- ( RSS_m / sigmasq ) - (m - 2 - 2 * (length(index) -1))
  C_p[i]
}

which.min(C_p)
C_p

index <- as.numeric(subsets$which[which.min(C_p),])
index <- which(index %in% c(1))
index

model6 <- bodyfat[, index]

summary(lm(bodyfat~., data = model6))

```

# Part (b)

```{r}

set.seed(10)

folds <- createFolds(bodyfat[,1], 10)
#folds

models <- list(model1,model2,model3,model4,model5,model6)
MSE <- matrix(0,6,10)

for(i in 1:6){
  for(j in 1:10){
    train <- models[[i]][-folds[[j]], ]
    test <- models[[i]][folds[[j]], ]
    fit <- lm(bodyfat ~., data  = train)
    predicted <- predict(fit, newdata = test, type = "response")
    mse <- mean((test[,1] - predicted)^2)
    mse
    MSE[i,j] <- mse

  }
}

MSE


rowSums(MSE)


finalmodel <- which.min(rowSums(MSE))

cat("Choose model", finalmodel, "\n")



```



# Part (c)

## Fit this model to the data.

```{r}

X <- models[[finalmodel]]

fit <- lm(bodyfat ~., data = X)
summary(fit)

```


## Perform regression diagnostics.  <---   further 

```{r}

#par(mfrow = c(2, 2))
plot(fit)

plot(fit, which = 4)
plot(fit, which = 6)

```


## Comment on the validity of the assumptions of the linear model.


1. The Residual VS Fitted plot does not suggests any violations of the assumptions of the linear model. 

2. The QQ plot of standardized residual shows that the standardized residuals are lighter tailed than normal(right most side). 

3. The square root of standardized residual vs fitted plot and standardized residual vs leverage plot both suggests 39th observation as a potential outlier. From each plots, potential influential points are 39, 216, 175 207.




## Identify influential observations and outliers.
```{r}


cooks <- data.frame(x = 1:nrow(bodyfat), cooks_distance = cooks.distance(fit))
cooks

ggplot(cooks, aes(x = x, y= cooks_distance, colour="red"))+ geom_point()+ geom_text(aes(label=x),hjust=0, vjust=0)

remove1 <- which(cooks[,2] %in% sort(cooks[,2], decreasing = T)[1:2])
remove1


influ <- data.frame(x = 1:nrow(bodyfat), influence = influence(fit)$hat)
influ
ggplot(influ, aes(x = x, y= influence, colour="red"))+ geom_point()+ geom_text(aes(label=x),hjust=0, vjust=0)


remove2 <- which(influ[,2] %in% sort(influ[,2], decreasing = T)[1:2])
remove2

```

The potential influential observations and outliers can be 39th and 175th observation as we can see in the plots above.


## Delete them if necessary and re-fit the model.

```{r}

X_refit <- X[-unique(c(remove1, remove2)), ]

refit <- lm(bodyfat ~., data = X_refit)
summary(refit)


cooks <- data.frame(x = 1:(nrow(bodyfat)-2), cooks_distance = cooks.distance(refit))
head(cooks)

ggplot(cooks, aes(x = x, y= cooks_distance, colour="red"))+ geom_point()+ geom_text(aes(label=x),hjust=0, vjust=0)



influ <- data.frame(x = 1:(nrow(bodyfat)-2), influence = influence(refit)$hat)
head(influ)
ggplot(influ, aes(x = x, y= influence, colour="red"))+ geom_point()+ geom_text(aes(label=x),hjust=0, vjust=0)



```


