---
title: 'STAT 151A: Lab 09, HW 3 Q1'
author: "Billy Fang"
date: "October 26, 2017"
output: html_document
---

```{r setup, include=FALSE, cache=T}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=F}
library(car)
library(ggplot2)
library(reshape2)
library(dplyr)
```

These are not meant to be canonical solutions for this homework question,
but rather a sketch of possible approaches for the purpose of discussion in section.

# Reading/cleaning data


```{r}
dat <- read.csv("thatch-ant.dat.txt")
head(dat)
dat <- dat[which(dat$Colony <= 6), ]
dat <- na.omit(dat)
dim(dat)
n <- dim(dat)[1]

unique(dat$Colony)
unique(dat$Distance)
unique(dat$Headwidth)
unique(dat$Headwidth..mm.)
# plot(dat$Headwidth, dat$Headwidth..mm.)

levels(dat$Size.class)
# re-order Size.class levels
dat$Size.class <- factor(dat$Size.class, levels(dat$Size.class)[c(1, 4, 5, 6, 3, 2)])


dat$Colony <- as.factor(dat$Colony)
dat$Distance <- as.factor(dat$Distance)
dat <- dat[, -5] # remove mm version of headwidth
```

# Exploratory data analysis




```{r}
p <- ggplot(dat)
p + geom_bar(aes(x=Colony)) + ggtitle("Histogram of colony size")
p + geom_boxplot(aes(x=Colony, y=Mass)) + ggtitle("Boxplots of mass by colony")
p + geom_boxplot(aes(x=Colony, y=Headwidth)) + ggtitle("Boxplots of head width by colony")
```

```{r}
# p + geom_bar(aes(x=Colony, fill=Size.class), position="dodge") + ggtitle("Histograms of size class by colony")
dat.size.colony <- melt(dcast(dat, Size.class ~ Colony, fun.aggregate=length), id.vars="Size.class", variable.name="Colony", value.name="Count")
ggplot(dat.size.colony, aes(x=Colony, y=Count, fill=Size.class)) + geom_bar(stat="identity", position=position_dodge(width=0.8), width=0.7) + geom_text(aes(label=Count), position=position_dodge(width=0.8), vjust=-0.5, size=2.5)


# p + geom_bar(aes(x=Colony, fill=Distance), position="dodge") + ggtitle("Histograms of distance by colony")
dat.distance.colony <- melt(dcast(dat, Distance ~ Colony, fun.aggregate=length), id.vars="Distance", variable.name="Colony", value.name="Count")
ggplot(dat.distance.colony, aes(x=Colony, y=Count, fill=Distance)) + geom_bar(stat="identity", position=position_dodge(width=0.8), width=0.7) + geom_text(aes(label=Count), position=position_dodge(width=0.8), vjust=-0.5, size=2.5)
```


```{r fig.width=10}
#coplot(Mass ~ Distance | Colony, data=dat, rows=1)
dat.tmp <- dat %>% group_by(Colony, Distance) %>% mutate(Mass.mean=mean(Mass)) %>% ungroup
ggplot(dat.tmp, aes(x=as.numeric(as.character(Distance)), y=Mass)) + geom_point(size=0.5) +
  # geom_smooth(method='loess', se=F, method.args=list(degree=1)) +
  geom_line(aes(y=Mass.mean), color='blue') +
  facet_wrap(~Colony, nrow=1) + theme(aspect.ratio=1) +
  ggtitle("Coplot of Mass ~ Distance | Colony") +
  xlab("Distance")
```

```{r}
p + geom_point(aes(x=Headwidth, y=Mass), size=0.5) + ggtitle("Mass vs. head width")
# p + geom_point(aes(x=Headwidth, y=Mass)) + facet_wrap(~Colony, nrow=1) + theme(aspect.ratio=1) + ggtitle("Coplot of Mass ~ Headwidth | Colony")
```


# Modeling

When you use `lm` with one categorical explanatory variable,
what dummy variables are used in the model? What happens when you remove the intercept
when calling `lm`? What is the answer to the above questions when there are more
than one categorical explanatory variable? Does the order in which you list the explanatory
variables matter?


- *Default `lm`.* If we do `lm(Mass ~ Colony + Distance + Headwidth + Size.class, data=dat)`,
then we are implicitly saying that ants in colony $c$, distance $d$, and size class $s$
follow the model
$$y = \beta_{\text{intercept}} + \beta_{\text{colony $c$}} + \beta_{\text{distance $d$}} + \beta_{\text{size class $s$}} + \beta_{\text{headwidth}} \cdot \text{headwidth} + \epsilon,$$
where
$$\beta_{\text{colony $1$}} = \beta_{\text{distance $0$}} = \beta_{\text{size class $<30$}} = 0.$$
(Remember the default behavior of `lm` for categorical variables. The only coefficients that appear are $\beta_{\text{intercept}}, \beta_{\text{colony $2$}}, \ldots, \beta_{\text{colony $6$}}, \beta_{\text{distance $1$}}, \ldots, \beta_{\text{distance $10$}},\beta_{\text{size class $30-34$}}, \ldots, \beta_{\text{size class $>43$}}$.)
- *No intercept `lm`.* If you do `lm(Mass ~ Colony + Distance + Headwidth + Size.class - 1, data=dat)`
as we do below,
then we are implicitly saying that ants in colony $c$, distance $d$, and size class $s$
follow the model
$$y = \beta_{\text{colony $c$}} + \beta_{\text{distance $d$}} + \beta_{\text{size class $s$}} + \beta_{\text{headwidth}} \cdot \text{headwidth} + \epsilon,$$
where
$$\beta_{\text{distance $0$}} = \beta_{\text{size class $<30$}} = 0.$$






The above sketch considers variations in how you (or `lm`) choose dummy variables
to encode categorical variables. Note that these changes (intercept or no intercept, etc.)
will not change the column space of the design matrix, so most things will remain the same
(fitted values, residuals, RSS, etc.). However, the coefficients $\widehat{\beta}$
will be different simply because the dummy variables represent different things
depending on how you encode the categorical variables.



```{r}
mods <- list()
mods <- c(mods, list(lm(Mass ~ Colony + Distance + Headwidth + Size.class - 1, data=dat)))
# mods <- c(mods, list(lm(sqrt(Mass) ~ Colony + Distance + Headwidth + Size.class - 1, data=dat)))
mods <- c(mods, list(lm(I(Mass^(0.4)) ~ Colony + Distance + Headwidth + Size.class - 1, data=dat)))
mods <- c(mods, list(lm(log(Mass) ~ Colony + Distance + Headwidth + Size.class - 1, data=dat)))
```

```{r fig.width=4, fig.height=4, fig.show='hold'}
mod <- mods[[1]]
print(summary(mod))
fit <- fitted(mod)
rstud <- rstudent(mod)
qplot(fit, rstud, size=I(0.5)) +
  geom_hline(yintercept=0) +
  xlab("Fitted values") +
  ylab("Studentized residuals") +
  geom_smooth(method='loess', se=F, method.args=list(degree=1)) +
  ggtitle("Studentized residuals vs. fitted values")

qplot(log(fit), log(abs(rstud)), size=I(0.5)) +
  geom_smooth(method='lm', se=F) + xlab("log fitted values") + ylab("log abs. stud. res.")

coef(lm(log(abs(rstud)) ~ log(fit)))
```


```{r fig.show="hold", fig.width=4}
for (mod in mods) {
  fitted <- fitted(mod)
  rstud <- rstudent(mod)
  # print(summary(mod))
  # dat.mod <- cbind(dat, Fitted=fitted(mod), Stud.Res=rstudent(mod))
  # print(
  #   qplot(qt((1:n) / (n+1), df.residual(mod)), sort(dat.mod$Stud.Res)) + geom_abline(slope=1)
  #   + xlab("t-distribution quantiles")
  #   + ylab("Sorted studentized residuals")
  #   + ggtitle(sprintf("Q-Q plot for %s model", colnames(mod$model)[1]))
  # )
  print(
    qplot(fitted, rstud, size=I(0.5))
    + geom_hline(yintercept=0)
    + xlab("Fitted values")
    + ylab("Studentized residuals")
    + geom_smooth(method='loess', se=F, method.args=list(degree=1))
    + ggtitle(sprintf("Studentized residuals vs. fitted values\nfor %s model", colnames(mod$model)[1]))
  )
  # print(ggplot(dat.mod, aes(x=log(Fitted), y=log(abs(Stud.Res)))) +
  #         geom_point(size=0.5) + xlab("log Fitted values") + ylab("log Studentized residuals") + geom_smooth(method='lm', se=F))
  # b <- coef(lm(log(abs(Stud.Res))~log(Fitted), data=dat.mod))[2]
  # print(b)
}
```



```{r fig.show="hold", fig.width=4}
for (mod in mods) {
  crPlot(mod, variable="Headwidth",
         main=sprintf("CR plot for head width\nfor %s model", colnames(mod$model)[1]))
}
```

```{r}
summary(mods[[3]])
```

```{r fig.show='hold', fig.width=2.5, fig.height=2.5}
mods.col <- list()
for (i in 1:length(unique(dat$Colony))) {
  mod.curr <- lm(log(Mass) ~ Distance + Headwidth + Size.class, data=dat, subset=(Colony == i))
  mods.col <- c(mods.col, list(mod.curr))
  print(summary(mod.curr))
  coefs.dist <- coef(mod.curr)[2:5]
  print(
    qplot(
      as.numeric(as.character(levels(dat$Distance))),
      c(0, coefs.dist), 
      geom="line") +
      xlab("Distance") + ylab("Offset from Distance zero") +
      ggtitle(sprintf("Mass differences by distance class,\n Colony %d", i)) +
      ylim(-0.2, 0.2)
  )
}
```

# Alternative models

## Transforming head width

```{r}
mod <- lm(Mass ~ . - 1, data=dat)
crPlot(mod, variable="Headwidth", main="CR plot for head width")
summary(mod)
```

```{r fig.width=8}
mod <- lm(Mass ~ . + I(Headwidth^2) - 1, data=dat)
crPlots(mod, terms=~Headwidth + I(Headwidth^2), main="CR plot for head width")
summary(mod)
```

```{r fig.show='hold', fig.width=2.5, fig.height=2.5}
mods.col <- list()
for (i in 1:length(unique(dat$Colony))) {
  mod.curr <- lm(log(Mass) ~ Distance + Headwidth + I(Headwidth^2) + Size.class, data=dat, subset=(Colony == i))
  mods.col <- c(mods.col, list(mod.curr))
  print(summary(mod.curr))
  coefs.dist <- coef(mod.curr)[2:5]
  print(
    qplot(
      as.numeric(as.character(levels(dat$Distance))),
      c(0, coefs.dist), 
      geom="line") +
      xlab("Distance") + ylab("Offset from Distance zero") +
      ggtitle(sprintf("Mass differences by distance class,\n Colony %d", i)) +
      ylim(-0.2, 0.2)
  )
}
```


## Distance as a continuous variable

```{r fig.width=8, fig.height=4}
dat$Distance <- as.numeric(as.character(dat$Distance))
mod <- lm(Mass ~ . - 1, data=dat)
summary(mod)
crPlots(mod, terms=~Headwidth+Distance)
```

```{r fig.width=8, fig.height=8}
mod <- lm(Mass ~ . + I(Headwidth^2) - 1, data=dat)
summary(mod)
crPlots(mod, terms=~Headwidth+I(Headwidth^2)+Distance, layout=c(2,2))
```

```{r fig.width=8, fig.height=8}
mod <- lm(Mass ~ . + I(Headwidth^2) + I(Distance^2) - 1, data=dat)
summary(mod)
crPlots(mod, terms=~Headwidth+I(Headwidth^2)+Distance+I(Distance^2), layout=c(2,2))
```


```{r fig.show='hold'}
mods.col <- list()
for (i in 1:length(unique(dat$Colony))) {
  mod.curr <- lm(log(Mass) ~ Distance + Headwidth + I(Headwidth^2) + Size.class, data=dat, subset=(Colony == i))
  mods.col <- c(mods.col, list(mod.curr))
  print(summary(mod.curr))
}

for (i in 1:length(unique(dat$Colony))) {
  print(coef(mods.col[[i]])[2])
}
```



