---
title: "Untitled"
author: "Jiyoon Clover Jeong"
date: "9/6/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

## R Markdown

Some mnath notation

\begin{enumerate}
  \item[(a)] Show that under these circumstances,
  \[
    B_{Y|X} = B_{X|Y} = r_{XY}
  \]
  where $B_{Y|X}$ is the least-squares slope for the simple regression of $Y$
  on $X$; $B_{X|Y}$ is the least-squares slope for the simple regression of
  $X$ on $Y$; and $r_{XY}$ is the correlation between the two variables. Show
  that the intercepts are also the same, $A_{Y|X} = A_{X|Y}$. 
\end{enumerate}


Code chunk

```{r prob6a, results='hide', echo=FALSE}
    #setwd(dir="~/Dropbox/Berkeley/STAT151A/")
    library(GGally)
    library(data.table)

    col_names <- c("mpg", "cylinders", "displacement", "horsepower", "weight",
                   "acceleration", "model_year", "origin", "car_name")

    cars_dt <- data.table(read.table(file="/Users/cloverjiyoon/2017Fall/Stat 151A/HW/HW1/auto-mpg.data.txt",
                                     header=FALSE,
                                     col.names=col_names))

    cars_dt[, cylinders := as.factor(cylinders)]
    cars_dt[, model_year := as.factor(model_year)]
    cars_dt[, origin := as.factor(origin)]
```

I chose to investigate the relationship between `mpg`,
  `displacement` (i.e. engine size) and `horsepower` because the
  latter two variables have a large correlation (`r cor(cars_dt$displacement, cars_dt$displacement)`), while also seeming to trend quite tightly with
  `mpg` (correlation of `r cor(cars_dt$mpg, cars_dt$displacement)`
  between `mpg` and `displacement` and of `r cor(cars_dt$mpg, cars_dt$displacement)` between `mpg` and `horsepower`). However,
  as the plots indicate, conditioning on either one does not induce any sort of
  independence between `mpg` and the other. On the other hand, there does
  seem to be only a slight amount of dependence between `mpg` and
  `acceleration` once you condition on `horsepower`.
 
