---
title: "STAT 151A Lab 3: What do all these numbers in `lm()` mean?"
author: "Billy Fang"
date: "September 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
bodyfat <- read.csv("bodyfat.csv")
head(bodyfat)
```

Let us model body fat percentage as a linear combination of weight, height, and abdomen.

```{r}
y <- bodyfat$bodyfat
X <- cbind(1, bodyfat[, c("Weight", "Height", "Abdomen")])
p <- 3
n <- length(y)
X <- as.matrix(X)
beta_hat <- solve(t(X) %*% X, t(X) %*% y)
# check beta_hat
beta_hat

y_hat <- X %*% beta_hat
e <- y - y_hat
y_bar <- mean(y)
RSS <- sum(e^2)
RegSS <- sum((y_hat - y_bar)^2)
TSS <- sum((y - y_bar)^2)

# check if RegSS + RSS = TSS
RegSS + RSS
TSS

# check R^2
R2 <- RegSS / TSS
R2
```

```{r}
fit <- lm(bodyfat ~ Weight + Height + Abdomen, data=bodyfat)
summary(fit)
```





#### $t$-test for individual coefficients

We showed that under the normal model $y = X\beta^* + \epsilon$
with $\epsilon \sim N_n(0, \sigma^2 I_n)$,
the mean of $\hat{\beta}$ is $\beta^*$, and
the covariance matrix of $\hat{\beta}$ is $\sigma^2 (X^\top X)^{-1}$.

Thus the variance of $\hat{\beta}_j$ is $\sigma^2 v_{jj}$,
where $v_{jj}$ is the $j$th diagonal element of $(X^\top X)^{-1}$.

Therefore, if we want to test the null hypothesis

$$H_0 : \beta_j^* = \beta_j^{(0)},$$
then the statistic

$$\frac{\hat{\beta}_j - \beta^*_j}{\sigma \sqrt{v_{jj}}},$$

is standard normal under the null hypothesis $H_0$.
If we knew $\sigma$, we could do a $Z$-test of this statistic
with the standard normal distribution, but we typically do not know $\sigma$.

In lecture notes and in lab today, we showed that

$$\frac{e^\top e}{n - p - 1} = \frac{\text{RSS}}{n - p - 1}$$
is an unbiased estimator of $\sigma^2$,
so the standard error of $\hat{\beta}_j$ is

$$\sqrt{\frac{e^\top e}{n - p - 1}} \sqrt{v_{jj}}.$$

It turns out (proof omitted) that
$$t = \frac{\hat{\beta}_j - \beta^*_j}{\sqrt{\frac{e^\top e}{n - p - 1}} \sqrt{v_{jj}}},$$
follows a $t$-distribution with $n - p - 1$ degrees of freedom.
Thus you can use this to test the null hypothesis. If $|t|$ is very large,
then we can reject the null hypothesis.



`lm()` does a $t$-test for $H_0 : \beta_j^* = 0$ for each $j$.
Let us manually do this for the `Weight` coefficient.
We can compute the standard error, $t$-statistic, and p-value.

```{r}
V <- solve(t(X) %*% X)
rse <- sqrt(RSS / (n - p - 1))
rse
se <- rse * sqrt(V[2, 2])
se
t <- beta_hat[2] / se
t
pval <- 2 * pt(t, n - p - 1)
pval
```

Let's check with `lm()`.

```{r}
summary(fit)
```


#### $F$-test for all slopes

`lm()` also tests the global null hypothesis that all slopes are zero.

$$H_0 : \beta^*_1 = \beta^*_2 = \cdots = \beta^*_k = 0.$$

Under this null hypothesis, the statistic
$$F_0 = \frac{\text{RegSS}/p}{\text{RSS}/(n-p-1)}$$
follows an $F$-distribution with $p$ and $n - p - 1$ degrees of freedom.
[See Section 6.2.1 of Fox.]
If $F_0$ is too large, we can reject the null hypothesis.

```{r}
Fstat <- (RegSS / p) / (RSS / (n - p - 1))
Fstat
pval <- 1 - pf(Fstat, p, n - p - 1)
pval
```

```{r}
summary(fit)
```

#### $F$-test for subset of slopes

Instead of testing for all slopes being zero, we can test for a subset being zero.

$$H_0 : \beta^*_1 = \beta^*_2 = \cdots = \beta^*_q = 0.$$

Let $\text{RSS}_1$, $\text{RegSS}_1$, and $\text{TSS}_1$ be for the full model,
and let $\text{RSS}_0$, $\text{RegSS}_0$, and $\text{TSS}_0$ be for the null model.

Why are the following true?
$$\text{RSS}_0 \ge \text{RSS}_1$$
$$\text{TSS}_0 = \text{TSS}_1$$
$$\text{RSS}_0 - \text{RSS}_1 = \text{RegSS}_1 - \text{RegSS}_0$$
$$\text{RegSS}_0 \le \text{RegSS}_1$$

The $F$-statistic to test this null hypothesis is
$$F = \frac{(\text{RegSS}_1 - \text{RegSS}_0) / q}{\text{RSS}_1 / (n - p - 1)}.$$
If this is too large, we can reject the null.

[Note that the "all slopes" $F$-statistic is a special case of this one: take $q = p$,
and note that $\text{RegSS}_0 = 0$.]

Let us test if both weight and abdomen are zero.
We have already fitted the full model, so we just need to find $\text{RegSS}_0$.

```{r}
q <- 2
fit0 <- lm(bodyfat ~ Height, data=bodyfat)
y_hat0 <- fitted(fit0)
RegSS0 <- sum((y_hat0 - y_bar)^2)
RegSS0
RegSS
Fstat <- ((RegSS - RegSS0) / q) / (RSS / (n - p - 1))
Fstat
pval <- 1 - pf(Fstat, q, n - p - 1)
pval
```

